\sffamily An EEG-based visual Brain-Computer Interface for gaze-free communication

### 2024-11-14 chapters/abstract.tex
Started processing comments MDV

Beyond gaze independence, the work also advances general \ac{erp} decoding
techniques by refining the structure of linear and multilinear estimators.
These methods improve classification accuracy across a range of \ac{bci}
conditions, particularly when training data are limited.
The introduction of structured regularization in both linear and multilinear
models enhances the interpretability of the classifiers and reduces training
time and computational complexity.
This allows for more efficient training and better performance on unseen data,
contributing to more reliable systems that are adaptable to various contexts and user needs.

@@

\paragraph{Keywords:} brain-computer interface, electroencephalography,
event-related potentials, covert visuospatial attention, (multi)linear
decoding, severe physical impairment

>>>>

Beyond gaze independence, the work also advances general \ac{erp} decoding
techniques by refining the structure of linear and multilinear estimators.
These methods improve classification accuracy across a range of \ac{bci}
conditions, particularly when training data are limited.
The introduction of structured regularization in both linear and multilinear
models enhances the interpretability of the classifiers and reduces training
time and computational complexity.
This allows for more efficient training and contributes to generalizability
allowing for more reliable systems that are adaptable to various contexts and
user needs.

@@

\paragraph{Keywords:} brain-computer interface, electroencephalography,
event-related potentials, covert visuospatial attention, (multi)linear
decoding, physical impairment

### 2024-11-14 chapters/abstract_fr.tex
Started processing comments MDV

\chapter*{R\'esum\'e}
Les individus ayant des handicaps moteurs sévères, tels que ceux atteints de
syndrome d'enfermement ou de dysfonctionnement oculomoteur, rencontrent des défis considérables
lors de l'utilisation d'interfaces cerveau-ordinateur (BCI)  traditionnelles.
Ces systèmes nécessitent que les utilisateurs dirigent leur regard vers des
cibles spécifiques, une tâche qui devient impraticable pour les personnes ayant un contrôle oculaire limité ou inexistant.
Ce travail vise \`a palier cette limitation en développant des méthodes de BCI
indépendantes du regard, en se concentrant sur l'amélioration de la attention visuo-spatiale (VSA)
cachée,
où les utilisateurs peuvent diriger leur attention vers une cible sans mouvements
oculaires correspondants.
Une contribution clé de ce travail est la compensation pour la compensation de
la variabilité de latence des potentiels évoqués (ERP), laquelle impacte
la performance de décodage dans la
VSA cachée.
En gérant cette variabilite, la méthode proposée améliore la précision de la communication
sans nécessiter de contrôle du regard, rendant les BCI plus utilisables pour les personnes ayant des handicaps moteurs.

Au-delà de l'indépendance du regard, le travail fait également progresser les
techniques générales de décodage ERP en affinant la structure des estimateurs
linéaires et multilinéaires.
Ces méthodes améliorent la précision de classification dans une gamme de
conditions BCI, en particulier lorsque les données d'entraînement sont limitées.
L'introduction de régularisation structurée dans les modèles linéaires et
multilinéaires améliore l'interprétabilité des classificateurs et réduit le temps d'entraînement ainsi que la complexité computationnelle.
Cela permet un entraînement plus efficace et une meilleure performance sur des données non vues, contribuant à des systèmes plus fiables qui s'adaptent à divers contextes et besoins des utilisateurs.

@@

\paragraph{Mots-clés :} interface cerveau-ordinateur, électroencéphalographie,potentiels évoqués
attnetion visuo-spatiale cachée, decodage (multi-)linéaire, patients lourdements handicapées

>>>>

\chapter*{R\'esum\'e}
Les individus ayant des handicaps moteurs sévères, tels que ceux atteints de
syndrome d'enfermement ou de dysfonctionnement oculomoteur, rencontrent des défis considérables
lors de l'utilisation d'interfaces cerveau-ordinateur (BCI)  traditionnelles.
Ces systèmes nécessitent que les utilisateurs dirigent leur regard vers des
cibles spécifiques, une tâche impraticable pour les personnes ayant un contrôle oculaire limité ou inexistant.
Ce travail vise \`a palier cette limitation par améliorer des méthodes de BCI
indépendantes du regard et la décodage de la attention visuo-spatiale (VSA) cachée, où les utilisateurs peuvent diriger leur attention vers une cible sans mouvements oculaires correspondants.
Une contribution clé de ce travail est la compensation pour la compensation de
la variabilité de latence des potentiels évoqués (ERP), laquelle impacte
la performance de décodage de VSA cachée.
En gérant cette variabilit\'e, la méthode proposée améliore la précision de la communication
sans nécessiter de contrôle du regard, rendant les BCI plus utilisables pour les personnes ayant des handicaps moteurs.

Au-delà de l'indépendance du regard, le travail fait également progresser les
techniques générales de décodage ERP en affinant la structure des estimateurs
linéaires et multilinéaires.
Ces méthodes améliorent la précision de classification dans une gamme de
conditions BCI, en particulier lorsque les données d'entraînement sont limitées.
L'introduction de régularisation structurée dans les modèles linéaires et
multilinéaires améliore l'interprétabilité des classificateurs et réduit le temps d'entraînement ainsi que la complexité computationnelle.
Cela permet un entraînement plus efficace et contribue à la generalisation, ce qui
permets le d\'evelopment des systèmes plus fiables qui s'adaptent à divers
contextes et besoins des utilisateurs.

@@

\paragraph{Mots-clés :}
interfaces neuronales directes,
\'electroenc\'ephalographie,
potentiels \'evoqu\'es,
attention visuo-spatiale cachée,
decodage (multi-)linéaire,
handicap\'es moteurs

### 2024-11-14 chapters/abstract_nl.tex
Started processing comments MDV

Personen met ernstige motorische beperkingen, zoals degenen met het
Locked-in Syndroom (LIS) of oog-motorische disfuncties, ondervinden
aanzienlijke uitdagingen bij het gebruik van traditionele visuele brein-computer interfaces (BCI's).
Deze systemen vereisen dat gebruikers hun blik richten op specifieke doelen, een taak die
onuitvoerbaar wordt voor personen met beperkte of
geen controle over oogbewegingen.
Dit werk richt zich op deze beperking door
BCI-methoden onafhankelijk van oogbewegingen te ontwikkelen, met een focus op
het verbeteren van de bedekte visuospatiële aandacht (VSA), waarbij gebruikers
hun aandacht kunnen richten op een doel zonder overeenkomstige oogbewegingen.
Een belangrijke bijdrage van dit werk is het compenseren van variabiliteit in
de timing van event-gerelateerde potentialen (ERP's), wat de decodeerprestaties
van bedekte VSA negatief beïnvloedt.
Door deze variabiliteit aan te pakken, verbetert de voorgestelde methode de
nauwkeurigheid van de communicatie zonder afhankelijkheid van oogcontrole,
waardoor BCI's bruikbaarder worden voor personen met motorische beperkingen van
de ogen.

Naast onafhankelijkheid van oogbewegingen, bevordert dit werk ook de algemene
ERP-decodeertechnieken door de structuur van lineaire en multilineaire
decodeertechnieken te verfijnen.
Deze methoden verbeteren de classificatienauwkeurigheid in een breed scala aan
BCI-condities, met name wanneer trainingsgegevens beperkt zijn.
De introductie van gestructureerde regularisatie in zowel lineaire als
multilineaire modellen verhoogt de interpretateerbaarheid van de classificators
en vermindert de trainingstijd en computationele complexiteit.
Dit maakt efficiëntere training mogelijk en zorgt voor betere prestaties bij
ongeziene data, wat bijdraagt aan betrouwbaardere systemen die zich kunnen
aanpassen aan verschillende contexten en gebruikersbehoeften.

@@

\paragraph{Trefwoorden:} brein-computer interface, electroencephalografie,
event-gerelateerde potentialen, bedekte visuospatiale attentie, (multi)lineaire
decodering, zware fysieke beperkingen

>>>>

Personen met ernstige motorische beperkingen, bijvoorbeeld het Locked-in
Syndroom (LIS) of oog-motorische disfunctie, ondervinden aanzienlijke
uitdagingen bij het gebruik van traditionele visuele brein-computer interfaces (BCI's).
Deze systemen vereisen dat gebruikers hun blik richten op specifieke doelen,
wat onuitvoerbaar wordt voor personen met beperkte of geen controle
over oogbewegingen.
Dit werk richt zich op deze beperking door BCI-methoden onafhankelijk van
oogbewegingen te ontwikkelen, met een focus op het verbeteren van de bedekte
visuospati\"ele aandacht (VSA), de  aandacht voor een doel zonder
overeenkomstige oogbewegingen.
Een belangrijke bijdrage van dit werk is het compenseren van variabiliteit in
de timing van event-gerelateerde potentialen (ERP's), wat de decodeerprestaties
van bedekte VSA negatief beïnvloedt.
Door deze variabiliteit aan te pakken, verbetert de voorgestelde methode de
nauwkeurigheid van communicatie onafhankelijkheid van oogcontrole,
waardoor BCI's bruikbaarder worden voor personen met deze motorische
beperkingen.

Naast onafhankelijkheid van oogbewegingen, verbetert dit werk ook algemene
ERP decodering door de structuur van lineaire en multilineaire technieken te verfijnen.
Deze methoden verbeteren de classificatienauwkeurigheid in een breed scala aan
BCI-condities, met name wanneer trainingsgegevens beperkt zijn.
De introductie van gestructureerde regularisatie in zowel lineaire als
multilineaire modellen verhoogt de interpretateerbaarheid van de classificators
en vermindert de trainingstijd en computationele complexiteit.
Dit maakt efficiëntere training mogelijk en draagt bij aan generalisatie, wat
de ontwikkeling van betrouwbaardere systemen toelaat die zich kunnen
aanpassen aan verschillende contexten en gebruikersbehoeften.

@@

\paragraph{Trefwoorden:}
brein-computer interface,
electroencephalografie,
event-gerelateerde potentialen,
bedekte visuospatiale aandacht,
(multi)lineaire decodering,
fysieke beperkingen

### 2024-11-14 chapters/bci.tex
Started processing comments MDV

\fullpagefig[topleft]{%
  \includegraphics[width=\fullpagefigwidth]{figures/bci/bci_loop.pdf}
}{%
  The \acs{bci} loop.
}{%
  The \ac{bci} loop.
  The user interacts with the \ac{bci} via a specific paradigm, in this case involving
  visual stimulation.
  Electrical neural signals are recorded, and neurophysiological features related to the
  paradigm are extracted.
  Using machine learning, a decision can then be made based on these features, which can
  be presented back to the user.
  In a closed-loop \ac{bci}, this feedback allows the user to adapt to the \ac{bci}.
}{fig:bci/loop}

@@

\fullpagefig[topleft]{%
  \includegraphics[width=\fullpagefigwidth]{figures/bci/recording_modalities.pdf}
}{%
  Various electrophysiology recording technologies.
}{%
	Various electrophysiology recording technologies and their respective position in a
	cross-section of the skull.
	There is a trade-off between invasiveness and spatial resolution.
}{fig:bci/recording}

@@

\fullpagefig{%
  \input{figures/bci/paradigms.tikz.tex}
}{%
  Overview of \acs{bci} paradigms.
}{%
  A \ac{bci} paradigm defines which brain activity will be used for control.
  They can be categorized in active, reactive and passive paradigms, depending on their
  reliance on external stimulation and active participation required from the user.
  Adapted from~\textcite{Muehl2014}.
}{fig:bci/paradigms}

@@

The epochs contain values representing the voltage of a specific channel relative to the
reference at a specific time relative to the stimulus.
In \ac{erp} classification, they are usually directly used as features, by concatenating
all channels in the matrix format epoch $\mat{X}$ into one feature vector such that
$\mat{x} = \text{vec}(\mat{X})$.

@@

Linear decoding techniques such as the \ac{lcmv} beamformer~\cite{Wittevrongel2016}
generally work well for classifying \ac{erp} epochs.
\Ac{lcmv} beamforming learns a weight vector $\mat{w}$ from a set of epochs that can be
multiplied with the feature vector of a new epoch to obtain a prediction score.
The weights $\mat{w}$ are obtained by minimizing the variance of the output
\begin{equation}
  \argmin_\mat{w}\mat{w}^\intercal\mat{C}
	\mat{w}^\intercal
\end{equation}
under the linear constraint
\begin{equation}
	\mat{a}^\intercal\mat{w} = 1
\end{equation}
with activation pattern $\mat{a}$ the difference between the average of the attended
epochs and the average of the non-attended epochs in the training data.
This is then solved by the method of Lagrange multipliers as
\begin{equation}
	\mat{w} =
  \frac{\mat{C}^{-1}\mat{a}^\intercal}
  {\mat{a}\mat{C}^{-1}\mat{a}^\intercal}
\end{equation}
with $\mat{C}$ the background covariance.

@@

\Ac{lda} is a more well-known method that is, in fact, nearly equivalent to
\ac{lcmv} beamforming~\cite{Treder2016} under certain conditions.
It assumes normally distributed data with a different mean per class but a common
covariance.
Similar to \ac{lcmv} beamforming, \ac{lda} learns a weight vector $\mat{w}$ that
satisfies
\begin{equation}
  \argmax_\mat{w}\frac{\mat{w}^\intercal\mat{S}_\text{b}\mat{w}}{\mat{w}^\intercal\mat{S}_\text{w}\mat{w}}
\end{equation}
The between-class scatter matrix $\mat{S}_\text{b}$ models the variability between the
expected responses, while the within-class scatter matrix $\mat{S}_\text{w}$ models the
expected noise.
The optimization criterion corresponds to finding a projection that maximizes the spread
between classes (signal) while minimizing the spread within classes (noise).
$\mat{w}$ can be obtained as the eigenvector of
$\mat{S}_\text{w}^{-1}\mat{S}_\text{b}$ corresponding to the largest eigenvalue.

@@

Another example of this is \ac{tlda}~\cite{Sosulski2022}, which assumes temporal
stationarity in the background noise to regularize the problem.
It does this by imposing a block-Toeplitz structure to the between-class covariance
matrix of \ac{lda}, such that every temporal covariance block corresponding to a channel
has the structure
\begin{equation}
  \mat{S}_\text{w}^{(c)} = \begin{bmatrix}
s_0 & s_1 & s_2 & \cdots & s_{k-1} \\
s_1 & s_0 & s_1 & \cdots & s_{k-2} \\
s_2 & s_1 & s_0 & \cdots & s_{k-3} \\
\vdots & \vdots & \vdots & \ddots & \vdots \\
s_{k-1} & s_{k-2} & s_{k-3} & \cdots & s_0
\end{bmatrix}
\end{equation}
This method is simple yet effective, reaching state-of-the-art decoding performance.

>>>>

\fullpagefig[topleft]{%
  \includegraphics[width=\fullpagefigwidth]{figures/bci/bci_loop.pdf}
}{%
  \caption[The \acs{bci} loop.]{%
    The \ac{bci} loop.
    The user interacts with the \ac{bci} via a specific paradigm, in this case involving
    visual stimulation.
    Electrical neural signals are recorded, and neurophysiological features related to the
    paradigm are extracted.
    Using machine learning, a decision can then be made based on these features, which can
    be presented back to the user.
    In a closed-loop \ac{bci}, this feedback allows the user to adapt to the \ac{bci}.
  }
  \label{fig:bci/loop}
}

@@

\fullpagefig[topleft]{%
  \includegraphics[width=\fullpagefigwidth]{figures/bci/recording_modalities.pdf}
}{%
  \caption[Various electrophysiology recording technologies.]{%
	  Various electrophysiology recording technologies and their respective position in a
	  cross-section of the skull.
	  There is a trade-off between invasiveness and spatial resolution.
  }
  \label{fig:bci/recording}
}

@@

\fullpagefig{%
  \input{figures/bci/paradigms.tikz.tex}
}{%
  \caption[Overview of \acs{bci} paradigms.]{%
    A \ac{bci} paradigm defines which brain activity will be used for control.
    They can be categorized in active, reactive and passive paradigms, depending on their
    reliance on external stimulation and active participation required from the user.
    Adapted from~\textcite{Muehl2014}.
  }
  \label{fig:bci/paradigms}
}

@@

The epochs contain values representing the voltage of a specific channel relative to the
reference at a specific time relative to the stimulus.
In \ac{erp} classification, they are usually directly used as features, by concatenating
all $C$ channels of length $S$ time samples in the matrix format epoch
$\mat{X}\in\mathbb{R}^{C\times S}$ into one feature vector
$\mat{x}\in\mathbb{R}^{CS}$ such that $\mat{x} = \text{vec}(\mat{X})$.

@@

Linear decoding techniques such as the \ac{lcmv} beamformer~\cite{Wittevrongel2016}
generally work well for classifying \ac{erp} epochs.
\Ac{lcmv} beamforming learns a weight vector $\mat{w}\in\mathbb{R}^{CS}$ from
a set of epochs that can be multiplied with the feature vector $\mat{x}$ of a
new epoch to obtain a prediction score.
The weights are obtained by minimizing the variance of the output
\begin{equation}
  \argmin_\mat{w}\mat{w}^\intercal\mat{C}
	\mat{w}^\intercal
\end{equation}
under the linear constraint
\begin{equation}
	\mat{a}^\intercal\mat{w} = 1
\end{equation}
with \emph{activation pattern} $\mat{a}\in\mathbb{R}^{CS}$ the difference between the
average of feature vectores of the attended and non-attended epochs in the
training data,
and $\mat{C}\in\mathbb{R}^{CS\times CS}$ the background covariance matrix which
can be estimated empirically from the training data.
This is then solved by the method of Lagrange multipliers as
\begin{equation}
	\mat{w} =
  \frac{\mat{C}^{-1}\mat{a}^\intercal}
  {\mat{a}\mat{C}^{-1}\mat{a}^\intercal}
\end{equation}

@@

\Ac{lda} is a more well-known method that is, in fact, nearly equivalent to
\ac{lcmv} beamforming~\cite{Treder2016} under certain conditions.
It assumes normally distributed data with a different mean per class but a common
covariance.
Similar to \ac{lcmv} beamforming, \ac{lda} learns a weight vector
$\mat{w}\in\mathbb{R}^{CS}$ which satisfies
\begin{equation}
  \argmax_\mat{w}\frac{\mat{w}^\intercal\mat{S}_\text{b}\mat{w}}{\mat{w}^\intercal\mat{S}_\text{w}\mat{w}}
\end{equation}
The between-class scatter matrix $\mat{S}_\text{b}\in\mathbb{R}^{CS\times CS}$ models the variability between the
expected responses, while the within-class scatter matrix $\mat{S}_\text{w}\in\mathbb{R}^{CS\times CS}$ models the
expected noise.
The optimization criterion corresponds to finding a projection that maximizes the spread
between classes (signal) while minimizing the spread within classes (noise).
$\mat{w}$ can be obtained as the eigenvector of
$\mat{S}_\text{w}^{-1}\mat{S}_\text{b}$ corresponding to the largest eigenvalue.

@@

Another example of this is \ac{tlda}~\cite{Sosulski2022}, which assumes temporal
stationarity in the background noise to regularize the problem.
It does this by imposing a block-Toeplitz structure to the between-class
scatter $\mat{S}_\text{b}$ matrix of \ac{lda}.
This means
\begin{equation}
  \mat{S}_\text{w} = \begin{bmatrix}
    \mat{S}_\text{w}^{(1,1)} & \mat{S}_\text{w}^{(1,2)} & \cdots & \mat{S}_\text{w}^{(1,C)}\\
    \mat{S}_\text{w}^{(2,1)} & \mat{S}_\text{w}^{(2,2)} & \cdots & \mat{S}_\text{w}^{(2,C)}\\
    \vdots             & \vdots             & \ddots & \vdots            \\
    \mat{S}_\text{w}^{(C,1)} & \mat{S}_\text{w}^{(C,2)} & \cdots & \mat{S}_\text{w}^{(C,C)}\\
  \end{bmatrix}
\end{equation}
consists of $C\times C$ temporal scatter matrices
$\mat{S}_\text{w}^{(c_1,c_2)}\in\mathbb{R}^{S\times S}$
corresponding to a pair of channels $(c_1,c_2)$, such
that every temporal scatter matrix has the Toeplitz structure
\begin{equation}
  \mat{S}_\text{w}^{(c_1,c_2)} = \begin{bmatrix}
    s_1^{(c_1,c_2)}  & s_2^{(c_1,c_2)}  & \cdots & s_S^{(c_1,c_2)}  \\
    s_2^{(c_1,c_2)}  & s_1^{(c_1,c_2)}  & \cdots & s_{S-1}^{(c_1,c_2)}  \\
    \vdots & \vdots &  \ddots & \vdots \\
    s_S^{(c_1,c_2)}  & s_{S-1}^{(c_1,c_2)}  & \cdots & s_1^{(c_1,c_2)}
\end{bmatrix}
\end{equation}
The structuring is performed by setting all elements to the average of their
corresponding block-subdiagonal and subsequently applying shrinkage
regularization.
This method is simple yet effective, reaching state-of-the-art decoding performance.

### 2024-11-14 chapters/bttda.tex
Started processing comments MDV

\subsection{Tensor \& tensor methods}

@@

Due to its heuristic nature, the greedy algorithm does not always result in the
set of ranks with the highest achievable performance.
In combination with the fact that the feature selection cutoff parameter was
fixed somewhat arbitrarily, it is clear that thorough
hyperparameter optimization could improve performance.
Furthermore, it is clear that our proposed model selection procedure does not
necessarily result in an optimal set of blocks that group coherent projections
within the same block, according to some desirable metric.
Examples of this are sparsity, pattern interpretability, minimal or maximal
between- or within-block feature correlation, decreasing discriminability, etc.
Finally, the completeness of the presented results is limited by the artificial
restriction in hyperparameter choice.
We imposed $(r_1=r_2=\ldots=r_k)$ that to reduce the computational demand of
the performed experiments.
In a sense, this goes against the proposition of increased model flexibility.
It might well be that \ac{bttda} in some cases offers little added value over the
Tucker-structured \ac{hoda}, when both are given free choice of rank.
Future efforts should focus on automatic parameter setting, e.g., using sparsity
or information criteria such as the ones used in Block-Term Tensor Regression~\cite{Faes2022}
or other statistical measures based on the model's application.

@@

Via an accompanying heuristic model selection procedure, \ac{bci}decoders using
\ac{bttda} feature extraction can significantly outperform decoders based on
\ac{hoda} and reach state-of-the-art decoding performance on event-related potential
problems (second-order tensors) and scores on par with or higher than \ac{hoda} in motor
imagery problems (third-order tensors).
Moving from the rigid Tucker tensor structure of \ac{hoda} to the more flexible
block-term structure shifts the problem from finding optimally constrained
multilinear projections to model and feature selection.

>>>>

\subsection{Tensors \& tensor methods}

@@

Due to its heuristic nature, the greedy algorithm does not always result in the
set of ranks with the highest achievable performance.
In combination with the fact that the feature selection cutoff parameter was
fixed somewhat arbitrarily, it is clear that thorough
hyperparameter optimization could improve performance.
Furthermore, it is clear that our proposed model selection procedure does not
necessarily result in an optimal set of blocks that group coherent projections
within the same block, according to some desirable metric.
Examples of this are sparsity, pattern interpretability, minimal or maximal
between- or within-block feature correlation, decreasing discriminability, etc.
Finally, the completeness of the presented results is limited by the artificial
restriction in hyperparameter choice.
We imposed $(r_1=r_2=\ldots=r_k)$ to reduce the computational demand of
the performed experiments.
In a sense, this goes against the proposition of increased model flexibility.
It might well be that \ac{bttda} in some cases offers little added value over the
Tucker-structured \ac{hoda}, when both are given free choice of rank.
Future efforts should focus on automatic parameter setting, e.g., using sparsity
or information criteria such as the ones used in Block-Term Tensor Regression~\cite{Faes2022}
or other statistical measures based on the model's application.

@@

Via an accompanying heuristic model selection procedure, \ac{bci} decoders using
\ac{bttda} feature extraction can significantly outperform decoders based on
\ac{hoda} and reach state-of-the-art decoding performance on event-related potential
problems (second-order tensors) and scores on par with or higher than \ac{hoda} in motor
imagery problems (third-order tensors).
Moving from the rigid Tucker tensor structure of \ac{hoda} to the more flexible
block-term structure shifts the problem from finding optimally constrained
multilinear projections to model and feature selection.
